A Hexagon Tile Map Algorithm for Displaying Spatial Data

### 2020-11-03 kobakian-cook.Rmd
Colours of centroid location

---
title: A Hexagon Tile Map Algorithm for Displaying Spatial Data
author:
  # see ?rjournal_article for more information
  - name: Stephanie Kobakian
    affiliation: Monash University
    address:
    - Department of Econometrics and Business Statistics
    email:  stephanie.kobakian@monash.edu 
  - name: Dianne Cook
    affiliation: Monash University
    address:
    - Department of Econometrics and Business Statistics
    email:  dicook@monash.edu 
abstract: >
  The algorithm presented creates a display that uses tessellated hexagons to represent a set of spatial polygons. It allocates these hexagon in a manner that preserves the spatial relationship of the geographic units, in light of their positions to points of interest. The display showcases spatial distributions, by emphasising the small geographical regions that are often difficult to locate on geographic maps. Spatial distributions have been presented on alternative representations of geography for many years. In modern times, interactivity and animation have allowed alternative displays to play a larger role. Alternative representations have been popularised by online news sites, and atlas websites with a focus on public consumption. Applications are increasingly widespread, especially in the areas of disease mapping, and election results.
preamble: |
  % Any extra LaTeX you need in the preamble
bibliography: kobakian-cook.bib
output: rticles::rjournal_article
editor_options: 
  chunk_output_type: console
---

@@

tas_centroids <- ggplot() + 
  geom_polygon(aes(x = long, y = lat, group = interaction(sa2_name_2011, polygon)), data = fort_tas, fill = "#D3D3D3", colour = "white", size = 0.01) + 
  geom_point(aes(x=longitude, y = latitude), size = 2, shape = 19, colour = "white",
             data = centroids) +
  scale_colour_identity() +
  theme_void() + 
  coord_equal()

tas_hexmap <- ggplot() +
  geom_polygon(aes(x = long, y = lat, group = interaction(sa2_name_2011, polygon)), data = fort_tas, fill = "#D3D3D3", colour = "white", size = 0.01) + 
  geom_polygon(aes(x = long, y = lat, group = hex_id, fill = SIR),
               data = fort_hex %>% filter(sa2_name_2011 %in% fort_tas$sa2_name_2011)) +
  scale_fill_identity() +
  theme_void() + 
  coord_equal()
```

@@

A grid is created to ensure tessellation between the hexagons that represent the geographic units on a hexagon tile map. The grid of possible hexagon locations is made using the `create_grid()` function. It uses the centroids, the hexagon size and the buffer distance.

@@

A set of longitude columns, and latitude rows are created to define the locations of the hexagons. The distance between each row and column is the size specified by `hex_size`. Equally spaced columns are created from the minimum longitude minus the buffer distance, up to the maximum longitude plus the buffer distance. Similarly, the rows are created from the latitude values and the buffer distance. A unique hexagon location is created from all intersections of the longitude columns and latitude rows. Figure \ref{fig:filterprocess}a) shows the hexagon grid after every second latitude row on the grid is shifted right, by half of the hexagon size.

@@

Not all of the grid points will be used, especially if islands result in a large grid space. To filter the grid for appropriate hexagon locations for allocation, the `create_grid` function will call the `create_buffer` function. It finds the grid points needed to best capture the set of centroids on a hexagon tile map. It also helps reduce the amount of time taken, as it decreases the amount of points considered for each centroid allocation.

@@

The hexagon grid points are kept only if they fall within the rolling average of the minimum and maximum centroid values after accounting for the buffer distance, for each row and column of the grid. Figure 5b displays remaining hexagon grid points after applying the buffer step. The sparsely populated South-West region of National Park has very few points available compared to the South-East region near the city of Hobart.

```{r filtergrid, fig.cap = "Each key step in the hexagon allocation process. The full hexagon grid in a) is created first. The buffer is applied and the remaining points are shown as larger, dark green circles, the potential hexagon points that were removed are shown as smaller, light green points. For each centroid in the set, c) shows the points left after the closest points are selected as potential locations. The relationship between Hobart (the cross) and the centroid (the yellow triangle is used to filter the potential locations within a wedge."}
g1 <- ggplot() + 
  geom_polygon(aes(x = long, y = lat, group = interaction(sa2_name_2011, polygon)), data = fort_tas, fill = "#D3D3D3", colour = "white", size = 0.01) + 
   theme_void() + 
  geom_point(aes(x = hex_long, y = hex_lat), colour = "#09661f", data = grid, size = 1) +
  coord_equal()

@@

g2 <- ggplot() + 
   geom_polygon(aes(x = long, y = lat, group = interaction(sa2_name_2011, polygon)), data = fort_tas, fill = "#D3D3D3", colour = "white", size = 0.01) +  
  theme_void() + coord_equal() +
  geom_point(aes(x = hex_long, y = hex_lat), colour = "#55b56c", data = full_grid, size = 0.5) +
  geom_point(aes(x = hex_long, y = hex_lat), colour = "#09661f", data = grid, size = 1)

@@

Allocation of all centroids takes place using the set of polygon centroids and the hexagon map grid. The points are arranged from the centroid closest to the focal point(s), to the furthest. Centroid allocation begins with the closest centroid to a focal point. This will preserve spatial relationships with the focal point, as the inner city areas are allocated first, they will be placed closest to the capital, and the areas that are further will then be accommodated. The set of possible hexagon grid points reduces by one after each centroid is allocated, then only the grid points that have not yet been selected are considered.

The possible hexagon locations to consider for a centroid are determined by the `hex_filter`. This is the maximum amount of hexagons between the centroid and the furthest considered hexagon. It is used to subset possible grid points to only those surrounding the polygon centroid within an appropriate range. A smaller distance will increase speed, but can decrease accuracy when width of the angle increases.

@@

The following example considers the first of the Statistical Areas at Level 2. Within the algorithm, these steps are repeated for each polygon.

@@

A box of possible hexagon locations around the centroid allows only the closest points that are not yet assigned to be considered. The corners of the box may not appear square if the buffer step has already removed unnecessary points from over the ocean.

@@

f_grid <- f_grid %>%
  group_by(id) %>%
  mutate(
    hex_lat_c = (hex_lat - flat),
    hex_long_c = (hex_long - flong)) %>%
  mutate(hyp = ((hex_lat_c^2) + (hex_long_c^2))^(1 / 2))
f_angle <- f_centroid %>%
      mutate(atan = atan2(latitude - focal_latitude, longitude - focal_longitude),
        angle = (atan * 180 / pi),
        pangle = ifelse(angle < 0, angle + 360, angle)) %>%
      pull(pangle)
f_grid <- f_grid %>%
  # create circle of radius: f_dist
  filter(hyp < f_dist) %>%
  mutate(
    # geosphere takes a long time
    angle = f_angle,
    angle_plus = (angle + angle_width) %% 360,
    angle_minus = (angle - angle_width) %% 360,
    atan = atan2(hex_lat_c, hex_long_c),
    hex_angle = (atan * 180 / pi),
    hex_angle = ifelse(hex_angle < 0, hex_angle + 360, hex_angle))

f1 <- ggplot() + 
  geom_polygon(aes(x = long, y = lat, group = interaction(sa2_name_2011, polygon)), data = fort_tas, fill = "#D3D3D3", colour = "white", size = 0.01) +
  theme_void() + coord_equal() +
  geom_point(aes(x = hex_long, y = hex_lat), colour = "#55b56c", data = grid, size = 0.5) +
  geom_point(aes(x = hex_long, y = hex_lat), colour = "#09661f", data = f_grid, size = 1) + 
  geom_point(aes(x=longitude, y = latitude), data = f_centroid, fill = "#ede187", shape = 25) + 
#Hobart
  geom_point(aes(x = focal_longitude, y = focal_latitude), data = f_centroid, colour = "#e9f2ed", size = 2, shape = 4)

The `width` parameter is used to take a slice of the remaining points. The width is the amount of degrees used on either side of the angle from the focal point to centroid location. This uses the angle from the closest capital city, to the current centroid as seen in Figure \ref{fig:filterprocess} . This allows the spatial relationship to be preserved, even when it is allocated to a hexagon that is further from the focal point then the original centroid location.

```{r filterprocess, fig.cap = "Filter for grid points within a square, then circular, distance for those closest to the centroid.", echo = FALSE, height = 12}
full_grid <- full_grid %>% filter(hex_long > bbox$min[1], hex_long < bbox$max[1],
                                  hex_lat > bbox$min[2], hex_lat < bbox$max[2])

@@

@@

f2 <- ggplot() + 
  geom_polygon(aes(x = long, y = lat, group = interaction(sa2_name_2011, polygon)), data = fort_tas %>% filter(lat < -40.3), fill = "#D3D3D3", colour = "white", size = 0.01) +
  geom_polygon(aes(x = long, y = lat, group = interaction(sa2_name_2011)), data = fort_allocated, fill = "#353b38", colour = "black", size = 0.01) +
  #Hobart
  geom_point(aes(x = focal_longitude, y = focal_latitude), data = f_centroid, colour = "#e9f2ed", size = 3, shape = 4) + 
  geom_point(aes(x=longitude, y = latitude), data = f_centroid, fill = "#ede187", shape = 25, size = 2) + 
 # theme_void() + coord_equal() +
  # possible choices to select
  geom_point(aes(x = hex_long, y = hex_lat), colour = "#09661f", data = f_grid, size = 1.5) +
  # selected hexagon
  geom_point(aes(x = hex_long, y = hex_lat), colour = "#09661f", fill = "#ede187", data = hex, size = 2.5, shape = 21)

@@

If no available hexagon grid point is found within the original filter distance and angle, the distance is expanded, only when a maximum distance is reached will the angle expand to accommodate more possible grid points.\
By default the angle filter to hexagon grid points that fall within the bounds of the angle from the focal point to the geographic centroid, plus and minus 30 degrees. This will increase if no points can be found within the `hex_filter` distance. The default angle of 30 was chosen to allow the algorithm to choose hexagons that best maintained the spatial relationship between the focal point and geographic centroid.

A complete hexagon tile map of Tasmania is created by applying the algorithm steps to each centroid. The hexagon tile map is used to visualise the Australian Cancer Atlas data [@TACA]. Two views of the same data are produced by filling according to the thyroid Cancer Standardised Incidence Rates (SIRs) downloaded from the Australian Cancer Atlas site. This small example in Figure \ref{fig:filterprocess} shows the group of blue areas in the Hobart CBD more prominently in the hexagon tile map (b). The small red areas visible in the choropleth map (a) along the north coast are much larger in the hexagon tile maps. The hexagon tile map shows less yellow, this no longer overwhelms the map space with the information regarding the rural areas.

When the algorithm is applied to the complete set of all Australian Statistical Areas at Level 2 it results in the hexagon tile map display shown in Figure \ref{fig:hexmap}. This display highlights the density of Australian capital cities, as it draws attention to the many communities in Sydney, Melbourne and Hobart. This display also highlights the disparity in the burden of thyroid cancer for females in the communities of these cities. There are several collections of red hexagons in Sydney that represent the communities with much higher rates of diagnosis than the Australian average. Brisbane also experiences higher than average rates of diagnosis, but has more orange than red, which represents much higher rates. The females in the cities of Adelaide and Perth  show much lower rates of diagnosis.

>>>>

---
title: A Hexagon Tile Map Algorithm for Displaying Spatial Data
author:
  # see ?rjournal_article for more information
  - name: Stephanie Kobakian
    affiliation: Monash University
    address:
    - Department of Econometrics and Business Statistics
    email:  stephanie.kobakian@monash.edu 
  - name: Dianne Cook
    affiliation: Monash University
    address:
    - Department of Econometrics and Business Statistics
    email:  dicook@monash.edu 
  - name: Earl Duncan
    affiliation: Queensland University of Technology
    address:
    - School of Mathematical Sciences
    email:  earl.duncan@qut.edu.au
abstract: >
  The algorithm presented creates a display that uses tessellated hexagons to represent a set of spatial polygons. It allocates these hexagon in a manner that preserves the spatial relationship of the geographic units, in light of their positions to points of interest. The display showcases spatial distributions, by emphasising the small geographical regions that are often difficult to locate on geographic maps. Spatial distributions have been presented on alternative representations of geography for many years. In modern times, interactivity and animation have allowed alternative displays to play a larger role. Alternative representations have been popularised by online news sites, and atlas websites with a focus on public consumption. Applications are increasingly widespread, especially in the areas of disease mapping, and election results.
preamble: |
  % Any extra LaTeX you need in the preamble
bibliography: kobakian-cook.bib
output: rticles::rjournal_article
editor_options: 
  chunk_output_type: console
---

@@

tas_centroids <- ggplot() + 
  geom_polygon(aes(x = long, y = lat, group = interaction(sa2_name_2011, polygon)), data = fort_tas, fill = "#D3D3D3", colour = "white", size = 0.1) + 
  geom_point(aes(x=longitude, y = latitude), size = 2, shape = 19, colour = "white",
             data = centroids) +
  scale_colour_identity() +
  theme_void() + 
  coord_equal()

tas_hexmap <- ggplot() +
  geom_polygon(aes(x = long, y = lat, group = interaction(sa2_name_2011, polygon)), data = fort_tas, fill = "#D3D3D3", colour = "white", size = 0.1) + 
  geom_polygon(aes(x = long, y = lat, group = hex_id, fill = SIR),
               data = fort_hex %>% filter(sa2_name_2011 %in% fort_tas$sa2_name_2011)) +
  scale_fill_identity() +
  theme_void() + 
  coord_equal()
```

@@

A grid is created to allow tessellation of the hexagons that represent the geographic units. For a hexagon tile map, the grid of possible hexagon locations is made using the `create_grid()` function. It uses the centroids, the hexagon size and the buffer distance. Figure \ref{fig:filterprocess}a) shows the initial locations of grid points created for Tasmania.

@@

A set of longitude columns, and latitude rows are created to define the locations of the hexagons. The distance between each row and column is the size specified by `hex_size`. Equally spaced columns are created from the minimum longitude minus the buffer distance, up to the maximum longitude plus the buffer distance. Similarly, the rows are created from the latitude values and the buffer distance. A unique hexagon location is created from all intersections of the longitude columns and latitude rows. Figure \ref{fig:filterprocess}a shows the hexagon grid after every second latitude row on the grid is shifted right, by half of the hexagon size.

@@

Not all of the grid points will be used, especially if islands result in a large grid space. To filter the grid for appropriate hexagon locations for allocation, the `create_grid` function will call the `create_buffer` function. It finds the grid points needed to best capture the set of centroids on a hexagon tile map. Reducing the buffer can help to decrease the amount of time taken, as there will be less points in the water to consider for each centroid allocation.

@@

The hexagon grid points are kept only if they fall within the rolling average of the minimum and maximum centroid values after accounting for the buffer distance, for each row and column of the grid. Figure \ref{fig:filterprocess}b displays remaining hexagon grid points after applying the buffer filter. The sparsely populated South-West region of National Park has fewer points available in the water compared to the South-East region near the city of Hobart.

```{r filtergrid, fig.cap = "Each key step in the hexagon allocation process. The full hexagon grid in a) is created first. The buffer is applied and the remaining points are shown as larger, dark green circles, the potential hexagon points that were removed are shown as smaller, light green points. For each centroid in the set, c) shows the points left after the closest points are selected as potential locations. The relationship between Hobart (the cross) and the centroid (the yellow triangle is used to filter the potential locations within a wedge."}
g1 <- ggplot() + 
  geom_polygon(aes(x = long, y = lat, group = interaction(sa2_name_2011, polygon)), data = fort_tas, fill = "#D3D3D3", colour = "white", size = 0.1) + 
   theme_void() + 
  geom_point(aes(x = hex_long, y = hex_lat), colour = "#09661f", data = grid, size = 1) +
  coord_equal()

@@

g2 <- ggplot() + 
   geom_polygon(aes(x = long, y = lat, group = interaction(sa2_name_2011, polygon)), data = fort_tas, fill = "#D3D3D3", colour = "white", size = 0.1) +  
  theme_void() + coord_equal() +
  geom_point(aes(x = hex_long, y = hex_lat), colour = "#55b56c", data = full_grid, size = 0.5) +
  geom_point(aes(x = hex_long, y = hex_lat), colour = "#09661f", data = grid, size = 1)

@@

Allocation of all centroids requires the set of polygon centroids and the hexagon map grid. The polygon centroids are ordered from the centroid closest to the focal point(s), to the furthest. This will preserve spatial relationships with the focal point, as the inner city areas are allocated first and placed closest to the capital, the areas that are further will then be accommodated. The following example considers the first of the Statistical Areas at Level 2. Within the algorithm, these steps are repeated for each polygon.

@@

@@

A selection of possible hexagon locations around the centroid allows only the closest points that are not yet assigned to be considered. The number of possible hexagon locations to consider for a centroid is determined by the `hex_filter`. This is the maximum amount of hexagons between the centroid and the furthest considered hexagon. It is used to subset possible grid points to only those surrounding the polygon centroid within an appropriate range. A smaller distance will increase speed, but can decrease accuracy when width of the angle increases.

@@

f_grid <- f_grid %>%
  group_by(id) %>%
  mutate(
    hex_lat_c = (hex_lat - flat),
    hex_long_c = (hex_long - flong)) %>%
  mutate(hyp = ((hex_lat_c^2) + (hex_long_c^2))^(1 / 2))
f_angle <- f_centroid %>%
      mutate(atan = atan2(latitude - focal_latitude, longitude - focal_longitude),
        angle = (atan * 180 / pi),
        pangle = ifelse(angle < 0, angle + 360, angle)) %>%
      pull(pangle)

f_grid <- f_grid %>%
  # create circle of radius: f_dist
  filter(hyp < f_dist) %>%
  mutate(
    # geosphere takes a long time
    angle = f_angle,
    angle_plus = (angle + angle_width) %% 360,
    angle_minus = (angle - angle_width) %% 360,
    atan = atan2(hex_lat_c, hex_long_c),
    hex_angle = (atan * 180 / pi),
    hex_angle = ifelse(hex_angle < 0, hex_angle + 360, hex_angle))

full_grid <- full_grid %>% filter(hex_long > bbox$min[1], hex_long < bbox$max[1],
                                  hex_lat > bbox$min[2], hex_lat < bbox$max[2])

@@

f1 <- ggplot() + 
  geom_polygon(aes(x = long, y = lat, group = interaction(sa2_name_2011, polygon)), data = fort_tas, fill = "#D3D3D3", colour = "white", size = 0.1) +
  theme_void() + coord_equal(ylim = c(-44, -40.1), xlim = c(144.5, 149)) +
  geom_point(aes(x = hex_long, y = hex_lat), colour = "#55b56c", data = grid, size = 0.5) +
  geom_point(aes(x = hex_long, y = hex_lat), colour = "#09661f", data = f_grid, size = 1) +
  # allocated areas
    geom_polygon(aes(x = long, y = lat, group = interaction(sa2_name_2011)), data = fort_allocated, fill = "#353b38", colour = "black", size = 0.01, alpha = 0.7) +
  #Hobart
  geom_point(aes(x = focal_longitude, y = focal_latitude), data = f_centroid, colour = "#e9f2ed", size = 4, shape = 3) + 
  geom_point(aes(x=longitude, y = latitude), data = f_centroid, fill = "#beaed4", shape = 25)

The `width` parameter is used to take a slice of the remaining points. The width is the amount of degrees used on either side of the angle from the focal point to centroid location. This uses the angle from the closest capital city, to the current centroid as seen in Figure \ref{fig:filterprocess}c. This allows the spatial relationship to be preserved, even when it is allocated to a hexagon that is further from the focal point then the original centroid location.

```{r filterprocess, fig.cap = "Filter for grid points within a square, then circular, distance for those closest to the centroid.", dpi = 600, echo = FALSE, height = 12}

@@

f2 <- ggplot() + 
  geom_polygon(aes(x = long, y = lat, group = interaction(sa2_name_2011, polygon)), data = fort_tas %>% filter(lat < -40.3), fill = "#D3D3D3", colour = "white", size = 0.1) +
  geom_polygon(aes(x = long, y = lat, group = interaction(sa2_name_2011)), data = fort_allocated, fill = "#353b38", colour = "black", size = 0.01, alpha = 0.7) +
  #Hobart
  geom_point(aes(x = focal_longitude, y = focal_latitude), data = f_centroid, colour = "#e9f2ed", size = 4, shape = 3) + 
  geom_point(aes(x=longitude, y = latitude), data = f_centroid, fill = "#beaed4", shape = 25, size = 2) + 
  theme_void() + coord_equal(xlim = c(146.9, 147.8), ylim = c(-43.35, -42.5)) +
  # possible choices to select
  geom_point(aes(x = hex_long, y = hex_lat), colour = "#09661f", data = f_grid, size = 2.5) +
  # selected hexagon
  geom_point(aes(x = hex_long, y = hex_lat), colour = "#09661f", fill = "#beaed4", data = hex, size = 3.5, shape = 21)

@@

If no available hexagon grid point is found within the original filter distance and angle, the distance is expanded, only when a maximum distance is reached will the angle expand to accommodate more possible grid points.
By default the angle filter for the hexagon grid points create plus and minus 30 degrees bounds of the angle from the focal point to the geographic centroid. This will increase if no points can be found within the `hex_filter` distance. The default angle of 30 was chosen to allow the algorithm to find hexagons that best maintained the spatial relationship between the focal point and geographic centroid.

<!-- Tasmania example has been removed -->
<!-- A complete hexagon tile map of Australia is created by applying the algorithm steps to each centroid. The hexagon tile map is used to visualise the Australian Cancer Atlas data [@TACA]. Two views of the same data are produced by filling according to the thyroid Cancer Standardised Incidence Rates (SIRs) downloaded from the Australian Cancer Atlas site. This small example in Figure \ref{fig:hexmap} shows the group of blue areas in the Hobart CBD more prominently in the hexagon tile map. The small red areas visible in the choropleth map (a) along the north coast are much larger in the hexagon tile maps. The hexagon tile map shows less yellow, this no longer overwhelms the map space with the information regarding the rural areas. -->
